---
title: "05_EDA"
author: "Dylan Lee"
date: "03/04/2020"
output: pdf_document
---

```{r libraries}
setwd("/Users/dylanlee/GitHub/r_data_science/05_EDA")
library(tidyverse) ## ggplot2, tibble, tidyr, readr, purrr, dplyr
library(nycflights13)
library(gapminder)
library(Lahman)
library(modelr)
```

Chapter 5:
Exploratory Data Analysis

Questions to consider:
1) What type of variation occurs within my variables?
2) What type of covariation occurs between my variables?

```{r Visualizing categorical data - bar chart}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
diamonds %>%
  count(cut)
```

```{r Visualizing continuous data - histogram}
ggplot(data = diamonds) + 
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
diamonds %>% 
  count(cut_width(carat, 0.5))

smaller <- diamonds %>%
  filter(carat < 3)
ggplot(data = smaller, mapping = aes(x = carat)) + 
  geom_histogram(binwidth = 0.1)

# overlaying multiple cuts
ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) + 
  geom_freqpoly(binwidth = 0.1)

ggplot(data = smaller, mapping = aes(x = carat)) + 
  geom_histogram(binwidth = 0.01)
```

```{r Old Faithful Geyser in Yellowstone National Park example}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25)
```

```{r Unusual Values}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) + 
  coord_cartesian(ylim = c(0,50))

(unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>%
  arrange(y))
```

It’s good practice to repeat your analysis with and without the outli‐ ers. If they have minimal effect on the results, and you can’t figure out why they’re there, it’s reasonable to replace them with missing values and move on. However, if they have a substantial effect on your results, you shouldn’t drop them without justification. You’ll need to figure out what caused them (e.g., a data entry error) and disclose that you removed them in your write-up.


```{r Missing Values}
diamonds2 <- diamonds %>%
  filter(between(y,3,20))

diamonds2 <- diamonds %>%
  mutate(y = ifelse(y < 3 | y > 20, NA, y))
ggplot(data = diamonds2, mapping = aes(x = x, y=y))  +
  geom_point()
ggplot(data = diamonds2, mapping = aes(x = x, y=y))  +
  geom_point(na.rm = TRUE)
```

```{r Missing values - NYC flights}
nycflights13::flights %>% 
  mutate(cancelled = is.na(dep_time),
         sched_hour = sched_dep_time %/% 100,
         sched_min = sched_dep_time %% 100,
         sched_dep_time = sched_hour + sched_min / 60) %>%
  ggplot(mapping = aes(sched_dep_time)) + 
  geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
```

```{r Visualize categorical and continuous variable}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)

ggplot(diamonds) + 
  geom_bar(mapping = aes(x = cut))

ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)

ggplot(data = diamonds, mapping = aes(x = cut, y = price)) + 
  geom_boxplot()
```

```{r reorder mpg example}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()

ggplot(data = mpg) + 
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))

ggplot(data = mpg) + 
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) + 
  coord_flip()
```

```{r two categorical variables}
ggplot(data = diamonds) + 
  geom_count(mapping = aes(x = cut, y = color))

diamonds %>% 
  count(color, cut)
diamonds %>% 
  count(color, cut) %>% 
  ggplot(mapping = aes(x = color, y = cut)) + 
  geom_tile(mapping = aes(fill = n))
```
If the categorical variables are unordered, you might want to use the seriation package to simultaneously reorder the rows and columns in order to more clearly reveal interesting patterns. For larger plots, you might want to try the d3heatmap or heatmaply packages, which create interactive plots.

```{r two continuous variables} 
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price)) # note, points overlap

ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price), alpha = 1/100)

ggplot(data = smaller) + 
  geom_bin2d(mapping = aes(x = carat, y = price))
ggplot(data = smaller) + 
  geom_hex(mapping = aes(x = carat, y = price))

ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)), varwidth = TRUE)
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_number(carat, 20)))

```

```{r Patterns and Models}
ggplot(data = faithful) + 
  geom_point(mapping = aes(x = eruptions, y = waiting))

mod <- lm(log(price) ~ log(carat), data = diamonds)
diamonds2 <- diamonds %>% 
  add_residuals(mod) %>%
  mutate(resid = exp(resid))
ggplot(data = diamonds2) + 
  geom_point(mapping = aes(x = carat, y = resid))
ggplot(data = diamonds2) + 
  geom_boxplot(mapping = aes(x = cut, y = resid))

ggplot(faithful, aes(eruptions)) + 
  geom_freqpoly(binwidth = 0.25)

diamonds %>% 
  count(cut, clarity) %>%
  ggplot(aes(clarity, cut, fill = n)) + 
  geom_tile()


```





















