---
title: "03_Data_Transform_dplyr"
author: "Dylan Lee"
date: "02/04/2020"
output: pdf_document
---

```{r libraries}
setwd("/Users/dylanlee/GitHub/r_data_science/01_Data_Viz")
library(tidyverse) ## ggplot2, tibble, tidyr, readr, purrr, dplyr
library(nycflights13)
library(gapminder)
library(Lahman)
```

Chapter 3:
Data Transformation with dplyr

```{r nyc flights data}
data(flights)
flights
```

• Pick observations by their values (filter()).
• Reorder the rows (arrange()).
• Pick variables by their names (select()).
• Create new variables with functions of existing variables (mutate()).
• Collapse many values down to a single summary (summa rize()).

```{r filter}
(dec25 <- filter(flights, month == 12, day == 25))
(jan1 <- filter(flights, month == 1, day == 1))


# & is “and,” | is “or,” and ! is “not.”

filter(flights, month == 11 | month == 12)
(nov_dec <- filter(flights, month %in% c(11,12)))

# filter() excludes FALSE and NA values, need to keep NA in condition if wanted to keep: is.na(x)
```

```{r arrange}
arrange(flights, year, month, day)

arrange(flights, desc(arr_delay))

# missing values are always sorted at the end
```

```{r select}
select(flights, year, month, day)

select(flights, year:day)

select(flights, -(year:day))

rename(flights, tail_num = tailnum)

select(flights, time_hour, air_time, everything()) # useful if have handful of variables want to move to the start of the df
```
• starts_with("abc") matches names that begin with “abc”. • ends_with("xyz") matches names that end with “xyz”.
• contains("ijk") matches names that contain “ijk”.
• matches("(.)\\1") selects variables that match a regular expression. This one matches any variables that contain repeated characters.

```{r mutate}
flights_sml <- select(flights,
                      year:day,
                      ends_with("delay"),
                      distance,
                      air_time)
mutate(flights_sml,
       gain = arr_delay - dep_delay,
       speed = distance / air_time * 60)

mutate(flights_sml,
       gain = arr_delay - dep_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours)

transmute(flights,
          gain = arr_delay - dep_delay,
          hours = air_time / 60,
          gain_per_hour = gain / hours)

transmute(flights,
          dep_time,
          hour = dep_time %/% 100,
          minute = dep_time %% 100)
```

```{r lead / lag}
(x <- 1:10)
lag(x)
lead(x)
x - lag(x)
(x != lag(x))
```

```{r cumulative and rolling aggregates}
x
cumsum(x)
cummean(x)
cumprod(x)
cummin(x)
cummax(x)
```

```{r ranking}
(y <- c(1,2,3,NA,3,4))
min_rank(y)
min_rank(desc(y))
```

```{r summarize}
summarize(flights, delay = mean(dep_delay, na.rm = TRUE))

by_day <- group_by(flights, year, month, day)
summarize(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

```{r combining multiple operations with the pipe}
by_dest <- group_by(flights, dest)
delay <- summarize(by_dest, count = n(),
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE))
delay <- filter(delay, count > 20, dest != "HNL")

ggplot(data = delay, mapping = aes(x = dist, y = delay)) + 
  geom_point(aes(size = count), alpha = 1/3) + 
  geom_smooth(se = FALSE)

##### OR #####

delays <- flights %>%
  group_by(dest) %>%
  summarize(count = n(),
            dist = mean(distance, na.rm =TRUE),
            delay = mean(arr_delay, na.rm = TRUE)) %>%
  filter(count > 20, dest != "HNL")
ggplot(data = delays, mapping = aes(x = dist, y = delay)) + 
  geom_point(aes(size = count), alpha = 1/3) + 
  geom_smooth(se = FALSE)
```

```{r missing values}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>%
  group_by(year,month,day) %>% 
  summarize(mean = mean(dep_delay))
```

```{r counts}
delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarize(delay = mean(arr_delay))
ggplot(data = delays, mapping = aes(x = delay)) + 
  geom_freqpoly(binwidth = 10)

delays <- not_cancelled %>%
  group_by(tailnum) %>% 
  summarize(delay = mean(arr_delay, na.rm = TRUE),
            n = n())
ggplot(data = delays, mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)

delays %>% 
  filter(n > 25) %>%
  ggplot(mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```

```{r baseball example}
(batting <- as_tibble(Lahman::Batting))

batters <- batting %>%
  group_by(playerID) %>%
  summarize(ba = sum(H, na.rm=TRUE) / sum(AB, na.rm=TRUE),
            ab = sum(AB, na.rm=TRUE))

batters %>%
  filter(ab > 100) %>%
  ggplot(mapping = aes(x = ab, y = ba)) + 
    geom_point() + 
    geom_smooth(se = FALSE)
  
```

```{r summary functions}
# measures of location
not_cancelled %>%
  group_by(year,month,day) %>%
  summarize(avg_delay1 = mean(arr_delay),
            avg_delay2 = mean(arr_delay[arr_delay > 0 ]))

# measures of spread
not_cancelled %>%
  group_by(dest) %>%
  summarize(distance_sd = sd(distance)) %>% 
  arrange(desc(distance_sd))

# measures of rank
not_cancelled %>%
  group_by(year,month,day) %>%
  summarize(first = min(dep_time),
            last = max(dep_time))

# measures of position
not_cancelled %>% 
  group_by(year,month,day) %>%
  summarize(first_dep = first(dep_time),
            last_dep = last(dep_time))

not_cancelled %>% 
  group_by(year,month,day) %>%
  mutate(r = min_rank(desc(dep_time))) %>%
  filter(r %in% range(r))

# counts: n(), sum(!is.na(x)), n_distinct(x)
not_cancelled %>%
  group_by(dest) %>%
  summarize(carriers = n_distinct(carrier)) %>%
  arrange(desc(carriers))

not_cancelled %>%
  count(dest)

not_cancelled %>%
  count(tailnum, wt = distance)

not_cancelled %>%
  group_by(tailnum) %>%
  summarize(n = sum(distance))

# counts and proportions of logical values
not_cancelled %>% 
  group_by(year,month,day) %>% 
  summarize(n_early = sum(dep_time < 500))

not_cancelled %>% 
  group_by(year,month,day) %>% 
  summarize(hour_perc = mean(arr_delay > 60))

```

```{r grouping by multiple variables}
daily <- group_by(flights, year, month, day)
(per_day <- summarize(daily, flights = n()))
(per_month <- summarize(per_day, flights = sum(flights)))
(per_year <- summarize(per_month, flights = sum(flights)))
```

```{r ungrouping}
daily %>% 
  ungroup() %>%
  summarize(flights = n())
```

```{r grouped mutates and filters}
flights_sml %>% 
  group_by(year,month,day) %>%
  filter(rank(desc(arr_delay)) < 10)

(popular_dests <- flights %>%
  group_by(dest) %>%
  filter(n() > 365))

popular_dests %>% 
  filter(arr_delay > 0) %>%
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>%
  select(year:day, dest, arr_delay, prop_delay)
```









